---
title: "Bayesian Data Analysis"
author: "Dr Niamh Cahill (she/her)"
date: "Bayesian Generalised Linear Models"
output:
  output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(DiagrammeR)
library(tidyverse)
library(rjags)
library(R2jags)
library(tidybayes)
library(bayesplot)
library(maps)
library(maptools)



fabric_faults <- read_csv("https://www.dropbox.com/s/czdos7zv76i2t8s/fabric_faults.csv?raw=1")

```



## Poisson Regression

\footnotesize

Suppose response $Y$ is a count response and takes values 0, 1, 2, 3, ...
We could model $Y$ as:

$$Y \sim \mbox{Poisson}(\mu), \;\;\; \mu > 0$$ 

  - Poisson is a discrete probability distribution that expresses the probability of a given number of events occurring in a fixed interval of time or space (distance, area, volume etc).

  - We assume these events occur with a known constant rate and independently of the time since the last event.

## Poisson Regression
\footnotesize

It follows that:

$$P[Y=k] = \displaystyle\frac{e^{-\mu}\mu^{k}}{k!}\;\;\; \mbox{for} \;\;\;k=0, 1, 2, ...$$ $$\mathbb{E}[Y] = \mu, \;\;\; \mbox{Var}[Y] = \mu.$$ 

 
The Poisson regression model is of the form:
$$\mbox{log}(\mu) = \beta_0 + \beta_1x_1 + ... + \beta_kx_k.$$

  - Poisson regression is a __generalized linear model__ with a log __link function__ that relates $\mathbb{E}[Y]$ to the __systematic component__, which is a linear function of explanatory variables used in the model.


## Why not use regular linear regression?

\footnotesize

  - So far, we discussed regression problems where the data were assumed to be normally distributed.

- That's not always an appropriate assumption, for example when dealing with count data, e.g. death counts due to lip cancer in Scotland;

  - Data not normally distributed.
  
  - Variance is not constant.

Note: for sufficiently large values of $\mu$, the normal distribution with mean $\mu$ and variance $\mu$ can be used as an approximation to the Poisson distribution. Thus for large counts we may be able to use linear regression. 



## Interpretation of parameter estimates
\footnotesize

Suppose $X_1$ increases from $x_1$ to $x_1+1$ and other $x$'s remain fixed. Then $\mu$ changes from 


$$\mu(x) = \mbox{exp}(\beta_0 + \beta_1 x_1 + ... + \beta_k x_k)$$ 
to 
\begin{eqnarray*}
\mu(x+1) &=& \mbox{exp}(\beta_0 + \beta_1 (x_1+1) + ... + \beta_k x_k) \\
&=& \mu(x)e^{\beta_1}
\end{eqnarray*}

Each unit increase in $x_1$ multiplies the mean response by $e^{\beta_1}$. 

If $\beta_1 = 0, e^{\beta_1}=1$ and the mean is unaffected by changes in $x_1$.

<!-- ## Poisson Regression: Introduction -->
<!-- \footnotesize -->
<!--   - So far, we discussed regression problems where the data were assumed to be normally distributed. -->

<!-- - That's not always an appropriate assumption, for example when dealing with -->

<!--     - count data, e.g. death counts due to lip cancer in Scotland; -->

<!-- - We will a discuss (hierarchical) poisson regression model to answer questions using these types of data.  -->

<!-- - Suppose that $y_1,y_2,\ldots,y_n$ are independent count observations. -->

<!-- - If $y_i\sim\mbox{Poisson}(\lambda)$ then -->

<!-- $${\mathbb P}\{y_i=y\}=\frac{\lambda^y \exp(-\lambda)}{y!}, \mbox{ where $\lambda>0$.}$$ -->
<!-- - Under this model,  -->
<!-- $${\mathbb E}(y)=\lambda \mbox{ and } {\mathbb V}\mbox{ar}(y)=\lambda.$$ -->
<!-- - So, the model cannot account for situations where $${\mathbb E}(y)\neq {\mathbb V}\mbox{ar}(y).$$ -->


## Count Data Example - Fabric Faults
\footnotesize

  -  Data: Numbers of faults found in 32 rolls of fabric produced in a particular factory (Hinde, J. (1982))    
  -  Predictor: the length of each roll. 
  
 
```{r, fig.width=4, fig.height=2.5}
ggplot(fabric_faults, aes(x = length, y = faults)) +
  geom_point()
```


## Model Specification

\footnotesize

  - A standard choice for count data is the Poisson data model:
  
$y_i|\lambda_i \sim$  Poisson($\lambda_i$)

-  We need to specify how the fabric length $x_i$ 
relates to the parameter of interest $\lambda_i$
- We can try a regression model, e.g.
log($\lambda_i$) = $\alpha + \beta (x_i - \bar{x})$
- This is the Poisson regression model, which is an
example of a generalized linear model, with

  - linear predictor  $\alpha + \beta (x_i - \bar{x})$ (linear in the parameters) and
  -  a log-link function, such that log(E($y_i$)) equals the linear predictor.



## JAGS Specification
\footnotesize
$y_i|\lambda_i \sim$  Poisson($\lambda_i$)

$log(\lambda_i) = \alpha + \beta (x_i - \bar{x})$

$\alpha \sim N(0, 10^2)$

$\beta \sim N (0,10^2)$

```{r, echo = TRUE}
poismodel = "
model{
	for( i in 1:n ) {
 #Poisson likelihood for observed counts
  y.i[i] ~ dpois(lambda.i[i])
  lambda.i[i] <- exp(alpha + beta*(x.i[i] - mean(x.i)))
	}
 #Prior distributions
  alpha ~ dnorm(0, 10^-2)
  beta ~ dnorm(0, 10^-2)
  
  for(i in 1:n){yrep[i] ~ dpois(lambda.i[i])}
}
"
```

```{r, echo = FALSE}
## Create the JAGS data list
jags.data <- list(y.i = fabric_faults$faults, 
                  x.i = fabric_faults$length, 
                  n = nrow(fabric_faults))

## Choose parameters to monitor
parnames <- c("lambda.i","alpha","beta","yrep")

mod <- jags(data = jags.data, 
            parameters.to.save=parnames, 
            model.file = textConnection(poismodel),
            n.iter = 10000,
            n.burnin = 2000,
            n.thin = 5,
            DIC = TRUE)

## Create output objects
m <- mod$BUGSoutput$sims.matrix

```

## Results: Parameters
\footnotesize

```{r, echo = FALSE}
par_summary <- m %>% 
                gather_rvars(alpha,beta) %>% 
                median_qi(.value)
par_summary
```

## Results: Model Fit

```{r,echo = FALSE, include = FALSE}
mu_ind <- 1:nrow(fabric_faults)
mu_samps <- m %>% spread_draws(lambda.i[mu_ind]) 
mu_summary <- mu_samps %>% 
                median_qi(.width = 0.95)

fabric_res <- fabric_faults %>% mutate(mu.i = mu_summary$lambda.i,
                          lower = mu_summary$.lower,
                          upper = mu_summary$.upper)

p <- ggplot(fabric_res, aes(x = length, y = faults)) +
  geom_point(alpha = 0.3) +
  geom_line(aes(x = length, y = mu.i)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  theme_bw() 
```

```{r, echo = FALSE}
p
```

## Posterior Predictive Check

```{r}
yrep <- mod$BUGSoutput$sims.list$yrep
y <- fabric_faults$faults
ppc_stat(y,yrep,  stat = "max",binwidth = 2)
```


