---
title: "Bayesian Data Analysis"
author: "Dr Niamh Cahill (she/her)"
date: "Bayesian Hierarcical Models - Lip Cancer Case Study"
output:
  output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(DiagrammeR)
library(tidyverse)
library(rjags)
library(R2jags)
library(tidybayes)
library(bayesplot)
library(maps)
library(maptools)



fabric_faults <- read_csv("https://www.dropbox.com/s/czdos7zv76i2t8s/fabric_faults.csv?raw=1")

```


## Example: Lip cancer study
\footnotesize

  - We will now examine incidence rates of lip cancer deaths in males in 56 counties of Scotland, registered in 1975–1980.
  
    - Wakefield, Jonathan. “Disease mapping and spatial regression with count data.” Biostatistics 8 2 (2007): 158-83.
  
  - The data consist of the observed and expected numbers of deaths
  
    - For each region the observed frequency of deaths $y_i$ was recorded for $i = 1, 2 \ldots, n$.
  
    - An expected count $e_i$ was calculated based on age-specific reference rates (based on all data for all counties combined) and the population-by-age distribution in the county.
 
 - A covariate measuring the proportion of the population engaged in agriculture, fishing, or forestry (AFF) is also available (exposure to sunlight is a risk factor for lip cancer)
    
  
## Lip cancer study: more info
\footnotesize
  - $SMR_i = y_i/e_i$ is referred to as the standardized morbidity ratio i.e., the ratio of the number of deaths observed in a population over a given period to the number that would be expected over the same period. 
  
- An estimate of the true (but unknown) relative risk is
denoted by $\theta_i$ in county $i$, which captures the relative difference in risk of dying for county $i$ as compared to the reference population. 

- Problem: the SMRs in some regions are based on small $e_is$ (i.e., for 22 counties $e_i < 5$) and the estimates are thus very uncertain.
  
- Question: How to estimate county-specific relative risks?
  
- Recall: that we also have the information on the percentage of the population
working outside (in agriculture etc.). How do we use this information to improve the relative risk estimates?

## Estimating region specific mortality

\footnotesize
  - As usual, we need to specify how the data relates to the parameter
of interest, so here how death counts $y_i$ relate to relative risk $\theta_i$.

- A standard choice for count data is the Poisson data model:
$y_i|\lambda_i \sim Poisson(\lambda_i)$, where $\lambda_i = E(y_i) = Var(y_i).$

- For the lip cancer data set, let's consider the Poisson model with
$\lambda_i = \theta_ie_i$, such that the mean death count is given by the expected
death count times the relative risk.

-  How to estimate $\theta_i$?

  - Three choices:
  
      1. Estimate $\theta_i$ for each county independently. 

      2. Assume $\theta_i$ is the same for all counties, $\theta_i$ = $\theta$: we would no longer
investigate whether there are differences in relative risks.

      3. Use a hierarchical model for the $\theta_i$'s to exchange information across
counties.

## Hierarchical model for estimating relative risks

\footnotesize

  - We will assume the data model: $y_i|\theta_i \sim Poisson(\theta_ie_i)$, $\theta_i > 0$.

  - A Gamma distribution results in full conditionals in closed form:
if we set 

$$\theta_i|\alpha, \beta \sim Gamma(\alpha, \beta)$$
with $E(\theta_i) = \alpha/\beta$ and $Var(\theta_i) = \alpha/\beta^2$, then we get a nice
interpretable form for the posterior  of $\theta_i$:

$$\theta_i|\alpha, \beta,y \sim Gamma(\alpha + y_i, \beta + e_i)$$

  - Think of $\alpha$ as "the prior number of death counts"

  - Think of $\beta$ as the "prior expected counts"

  - The posterior mean will be a weighted sum of the prior mean and the observed rel. risks

$\frac{\alpha + y_i}{\beta + e_i} = w_i \frac{\alpha}{\beta} + (1-w_i) \frac{y_i}{e_i}$

where the weight (or shrinkage factor) $w_i = \frac{\beta}{\beta + e_i}$

## JAGS Model

\footnotesize

```{r, echo = TRUE}
bhpoismodel = "
model{
for(i in 1:n) {
# Poisson likelihood (data model)
y.i[i] ~ dpois(lambda.i[i])
lambda.i[i] <- theta.i[i]*e.i[i]

# prior for theta
theta.i[i] ~ dgamma(alpha,beta)
# Shrinkage factor
w.i[i] <- beta/(beta + expect.i[i])
}
# Prior distributions for hyper parameters
alpha ~ dexp(0.1)
beta ~ dexp(0.1)
# store group mean and across-group variance
mtheta <- alpha/beta
sdtheta <- sqrt(alpha/(beta^2))
}
"

```

```{r, echo = FALSE, include=FALSE}

library(SpatialEpi)
data(scotland)

lip_cancer_dat <- scotland$data
lip_cancer_dat <- lip_cancer_dat %>% mutate(SMR = cases/expected)

bhpoismodel = "
model{
for(i in 1:n) {
# Poisson likelihood (data model)
y.i[i] ~ dpois(lambda.i[i])
lambda.i[i] <- theta.i[i]*e.i[i]

# prior for theta
theta.i[i] ~ dgamma(alpha,beta)
# Shrinkage factor
w.i[i] <- beta/(beta + e.i[i])
}
# Prior distributions for hyper parameters
alpha ~ dexp(0.1)
beta ~ dexp(0.1)
# store group mean and across-group variance
mtheta <- alpha/beta
sdtheta <- sqrt(alpha/(beta^2))
}
"
jags.data <- list(y.i = lip_cancer_dat$cases,
                  e.i = lip_cancer_dat$expected,
                  n = nrow(lip_cancer_dat))
  
parnames <- c("alpha","beta","mtheta","sdtheta","w.i","theta.i")
  
mod <- jags(data = jags.data, 
              parameters.to.save = parnames, 
              model.file = textConnection(bhpoismodel),
              n.iter = 6000,
              n.burnin = 2000,
              n.thin = 4)

m <- mod$BUGSoutput$sims.matrix
```

## Let's explore the impact of the shrinkage 

\footnotesize
```{r, fig.width=9, fig.height=5}
index <- 1:nrow(lip_cancer_dat)
par_summary <- m %>% 
                gather_rvars(w.i[index],theta.i[index]) %>% 
                median_qi(.value)

par_wide <- par_summary %>% 
              select(index, .variable, .value) %>% 
              pivot_wider(id_cols = index, 
                          values_from = .value,
                          names_from = .variable) %>% 
              mutate(SMR = lip_cancer_dat$SMR,
                     expected = ifelse(lip_cancer_dat$expected<5,"< 5",">/= 5"))
prior_mean <- mod$BUGSoutput$mean$alpha/ mod$BUGSoutput$mean$beta

p1 <- ggplot(par_wide, aes(x = SMR, y = theta.i)) +
  geom_point(aes(colour = expected)) +
  geom_line(aes(x = SMR, y = SMR)) +
  geom_segment(aes(x = 1.3, y = prior_mean, xend = 0, yend = prior_mean),
                  arrow = arrow(length = unit(0.6, "cm"))) +
  annotate("text", x = 0.7, y = prior_mean + 0.4, label = "prior mean = 1.4")

p2 <- ggplot(par_wide, aes(x = lip_cancer_dat$expected, y = w.i)) +
  geom_point(aes(colour = expected)) +
  ylab("Shrinkage factor (w.i)") +
  xlab("Expected")

ggpubr::ggarrange(p1,p2, common.legend = T)

```

This behavior illustrates both the potential benefits and hazards of smoothing; 

  - the estimates of $\theta_i$ will be less variable than the observed SMRs, 
  - but an outlying estimate that is not based on a large expected number will be shrunk, and we may miss an important excess.

## Let's visualise some results

Raw and smoothed estimates in 56 counties of Scotland. 

```{r}
shapefile <- scotland$spatial.polygon

#Reshape for ggplot2 using the Broom package
mapdata <- broom::tidy(shapefile) #This might take a few minutes
lip_cancer_dat <- lip_cancer_dat %>% 
                      mutate(id =  mapdata$id %>% unique(),
                      SMR = cases/expected,
                      theta.i = mod$BUGSoutput$mean$theta.i)

#Join mydata with mapdata
df <- plyr::join(mapdata, lip_cancer_dat, by="id")


#Create the heatmap using the ggplot2 package
gg1 <- ggplot() + geom_polygon(data = df, aes(x = long, y = lat, group = group, fill = SMR), color = "#FFFFFF", size = 0.25)+ 
  scale_fill_gradient(low = "yellow", high = "red", na.value = "white",limits = c(0,7))+
  coord_fixed(1)+
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.title.x=element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  theme(axis.title.y=element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())

#Create the heatmap using the ggplot2 package
gg2 <- ggplot() + geom_polygon(data = df, aes(x = long, y = lat, group = group, fill = theta.i), color = "#FFFFFF", size = 0.25)+ 
  scale_fill_gradient(low = "yellow", high = "red", na.value = "white",limits = c(0,7))+
  coord_fixed(1)+
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.title.x=element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  theme(axis.title.y=element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggpubr::ggarrange(gg1 + ggtitle("Raw SMR Estimates") ,gg2 + ggtitle("Model Estimates"))

```

## Add in predictors for estimating the SMR

\footnotesize

  - We have information on the the proportion of the population engaged in agriculture, fishing, or forestry (AFF) for each county ($x_i$)
  
  - A possible set-up:
  
$$y_i|\lambda_i \sim Poisson(\lambda_i)$$

$$\lambda_i = \theta_i e_i$$
$$log(\theta_i) = \alpha_i + \beta(x_i - \bar{x})$$
$$\alpha_i|\mu_{\alpha},\sigma_{\alpha} \sim N(\mu_{\alpha},\sigma_{\alpha})$$

  - Note: We switched to using normal distributions for the log-transformed county-specific intercepts.

## Add in predictors for estimating the SMR

\footnotesize

```{r, echo=TRUE}
bhpoismodel2 = "
model{
for(i in 1:n) {
# Poisson likelihood (data model)
y.i[i] ~ dpois(lambda.i[i])
lambda.i[i] <- theta.i[i]*e.i[i]

# prior for theta
log(theta.i[i]) <- ltheta.i[i]

ltheta.i[i] <- alpha.i[i] + beta*(x.i[i]-mean(x.i))
alpha.i[i] ~ dnorm(mu_alpha, sigma_alpha^-2)

}
# Prior distributions for hyper parameters
mu_alpha ~ dnorm(0,0.01)
beta ~ dnorm(0,0.01)
sigma_alpha ~ dt(0,2^-2,1)T(0,)
}
"

```

```{r, echo = FALSE, include=FALSE}

bhpoismodel2 = "
model{
for(i in 1:n) {
# Poisson likelihood (data model)
y.i[i] ~ dpois(lambda.i[i])
lambda.i[i] <- theta.i[i]*e.i[i]

# prior for theta
log(theta.i[i]) <- ltheta.i[i]

ltheta.i[i] <- alpha.i[i] + beta*(x.i[i]-mean(x.i))
alpha.i[i] ~ dnorm(mu_alpha, sigma_alpha^-2)


}
# Prior distributions for hyper parameters
mu_alpha ~ dnorm(0,0.01)
beta ~ dnorm(0,0.01)
sigma_alpha ~ dt(0,2^-2,1)T(0,)
}
"

jags.data <- list(y.i = lip_cancer_dat$cases,
                  e.i = lip_cancer_dat$expected,
                  x.i = lip_cancer_dat$AFF,
                  n = nrow(lip_cancer_dat))
  
parnames <- c("alpha.i","beta","mu_alpha","sigma_alpha","theta.i")
  
mod <- jags(data = jags.data, 
              parameters.to.save = parnames, 
              model.file = textConnection(bhpoismodel2),
              n.iter = 10000,
              n.burnin = 2000,
              n.thin = 4)

m <- mod$BUGSoutput$sims.matrix
```


## Let's look at $exp(\alpha_i)$ 

\footnotesize

  - Boxplots are shown for $exp(\alpha_i)$ which is the expected SMR for counties where $\sim 9 \%$ work outside.

  - The solid black line is $exp(\mu_{\alpha})$ which is the expected SMR across all counties when $\sim 9\%$ work outside.

```{r, warning=FALSE, fig.width=6,fig.height=4}

alpha_ind <- 1:jags.data$n
alpha_samps <- m %>% spread_draws(alpha.i[alpha_ind]) %>% 
    dplyr::select(alpha_ind, alpha.i) %>% dplyr::ungroup() %>% 
    dplyr::mutate(alpha_ind = factor(alpha_ind, labels = lip_cancer_dat$county.names),
                  exp_alpha.i = exp(alpha.i))

colour_e <- ifelse(lip_cancer_dat$expected<5,"red","yellow")

ggplot(alpha_samps, aes(x = alpha_ind, y = exp_alpha.i)) +
  geom_boxplot(colour = colour_e,outlier.shape = NA) +
  geom_hline(yintercept = exp(mod$BUGSoutput$mean$mu_alpha))+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  xlab("") +
  ylab("exp(alpha.i)")

```

## Compare $exp(\alpha_i)$ for an independent vs hierarchical model

\footnotesize

  - The comparison below is for 3 counties that have expected $<5$
  
  - Note the difference in the uncertainty!!


```{r, echo = FALSE, include = FALSE}
bhpoismodel_alt = "
model{
for(i in 1:n) {
# Poisson likelihood (data model)
y.i[i] ~ dpois(lambda.i[i])
lambda.i[i] <- theta.i[i]*e.i[i]

# prior for theta
log(theta.i[i]) <- ltheta.i[i]

ltheta.i[i] <- alpha.i[i] + beta*(x.i[i]-mean(x.i))
alpha.i[i] ~ dnorm(0,0.01)

}
# Prior distributions for hyper parameters
beta ~ dnorm(0,0.01)
}
"

parnames <- c("alpha.i","beta","theta.i")
  
mod_alt <- jags(data = jags.data, 
              parameters.to.save = parnames, 
              model.file = textConnection(bhpoismodel_alt),
              n.iter = 10000,
              n.burnin = 2000,
              n.thin = 4)

m_alt <- mod_alt$BUGSoutput$sims.matrix

alpha_samps_alt <- m_alt %>% spread_draws(alpha.i[alpha_ind]) %>% 
    dplyr::select(alpha_ind, alpha.i) %>% dplyr::ungroup() %>% 
    dplyr::mutate(alpha_ind = factor(alpha_ind, labels = lip_cancer_dat$county.names),
                  exp_alpha.i = exp(alpha.i))

alpha_samps3<- alpha_samps %>% filter(alpha_ind %in% lip_cancer_dat$county.names[c(1,4,6)]) %>% mutate(model = "hierarcical")

alpha_samps_alt3 <- alpha_samps_alt %>% filter(alpha_ind %in% lip_cancer_dat$county.names[c(1,4,6)]) %>% mutate(model = "independent")

alpha_dat3 <- full_join(alpha_samps3,alpha_samps_alt3)

scales <- list(
  scale_x_continuous(limits = c(0,3)),
  scale_x_continuous(limits = c(0, 20))
)

p <- ggplot(alpha_dat3, aes(y = alpha_ind, x = exp_alpha.i)) +
  geom_boxplot(outlier.shape = NA) +
  #geom_hline(yintercept = exp(mod$BUGSoutput$mean$mu_alpha))+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  xlab("") +
  ylab("exp(alpha.i)") +
  xlim(0,16) +
  geom_vline(xintercept = exp(mod$BUGSoutput$mean$mu_alpha)) +
  facet_wrap(~model, scales = "free_x") 
```

```{r, echo = FALSE, warning = FALSE, fig.width=6, fig.height=3}
p 
```




## Let's look at $\theta_i$

```{r}
theta_ind <- 1:jags.data$n
theta_samps <- m %>% spread_draws(theta.i[theta_ind]) %>% 
    dplyr::select(theta_ind, theta.i) %>% dplyr::ungroup() %>% 
    dplyr::mutate(theta_ind = factor(theta_ind, labels = lip_cancer_dat$county.names))

colour_e <- ifelse(lip_cancer_dat$expected<5,"red","yellow")

ggplot(theta_samps, aes(x = theta_ind, y = theta.i)) +
  geom_boxplot(colour = colour_e,outlier.shape = NA) +
  geom_point(data = lip_cancer_dat, aes(x = theta_ind, y = SMR, colour = "observed SMR")) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  xlab("") +
  labs(colour = "") +
  ylab("theta.i")

```

## Let's visualise on a map

\footnotesize

```{r, fig.width=12,fig.height=6}
shapefile <- scotland$spatial.polygon

#Reshape for ggplot2 using the Broom package
mapdata <- broom::tidy(shapefile) #This might take a few minutes
lip_cancer_dat <- lip_cancer_dat %>% 
                      mutate(id =  mapdata$id %>% unique(),
                      theta.i = mod$BUGSoutput$mean$theta.i)

#Join mydata with mapdata
df <- plyr::join(mapdata, lip_cancer_dat, by="id")

#Create the heatmap using the ggplot2 package
gg3 <- ggplot() + geom_polygon(data = df, aes(x = long, y = lat, group = group, fill = theta.i), color = "#FFFFFF", size = 0.25)+ 
  scale_fill_gradient(low = "yellow", high = "red", na.value = "white",limits = c(0,7))+
  coord_fixed(1)+
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.title.x=element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  theme(axis.title.y=element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  ggtitle("with predictor")

#Create the heatmap using the ggplot2 package
gg4 <- ggplot() + geom_polygon(data = df, aes(x = long, y = lat, group = group, fill = AFF), color = "#FFFFFF", size = 0.25)+ 
  scale_fill_gradient(low = "green", high = "blue", na.value = "white",limits = c(0,0.3))+
  coord_fixed(1)+
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.title.x=element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  theme(axis.title.y=element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank()) 


ggpubr::ggarrange(gg1 + ggtitle("Raw SMR"),gg2 + ggtitle("w/o predictor"),gg3, gg4)

```
