---
title: "Bayesian Analysis"
author: "Dr Niamh Cahill (she/her)"
date: "Single Parameter Normal Model"
output:
  output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(rstanarm)
data(kidiq)
```

## Example: Cognitive Test Scores
\footnotesize
Data (y) are available on the cognitive test scores of three- and four-year-old children in the USA. The sample contains 434 observations.

```{r, fig.height=2.5, fig.width=4}
ggplot(kidiq, aes(x = kid_score)) +
  geom_histogram(bins = 30,colour = "blue")
```


## Normal distribution with known variance

\footnotesize
We will assume a normal model for the data where $y_i|\mu,\sigma^2 \sim N(\mu, \sigma^2)$. Assume $\sigma^2$ is known where $\sigma = 20.4$


 - Specify the likelihood for $\mu$
  
  
$p(y|\mu, \sigma^2) = \prod_{i=1}^{n} \frac{1}{\sqrt {2\pi\sigma^2}}exp \bigg(-\frac{1}{2\sigma^2}(y_i - \mu)^2\bigg)$


  - Specify a prior for $\mu$

 $\mu \sim N(\mu_0, \sigma^2_{_0})$

  - Use Bayes' rule to obtain the posterior distribution

$p(\mu|y) \propto p(y|\mu)p(\mu)$


  - As it turns out, the posterior is also a normal distribution

 $$\mu|y \sim N \bigg(\frac{n\bar{y}/\sigma^2 + \mu_0/\sigma^2_{0}}{n/\sigma^2 + 1/\sigma^2_{0}}, {\frac{1}{n/\sigma^2 + 1/\sigma^2_{0}}}\bigg)$$

## Prior vs Posterior for $\mu$

\footnotesize

  - Choose prior mean and variance, e.g., $\mu_0 = 80$, $\sigma_{0} = 10$
  
```{r,  fig.width=7, fig.height=2}
y <- kidiq$kid_score
n <- length(y)
#-------------------
# exact solutions
# prior and posterior for mu, conditional on sigma (assuming sigma = sd.y)
sd.y<- 20.4

# prior parameters
mu0 <- 80
sigma0 <- 10

# Posterior of mu, assuming sigma = sd.y (sample SD)
mupost.mean = (n*mean(y)/(sd.y^2) + mu0/(sigma0^2))/(n/(sd.y^2) + 1/(sigma0^2))
mupost.sd = sqrt(1/(n/(sd.y^2) + 1/(sigma0^2)))

# Credible interval
lwr <- qnorm(0.025, mean = mupost.mean, sd = mupost.sd)
upr <- qnorm(0.975, mean = mupost.mean, sd = mupost.sd)

# set up for plotting
mu_grid <- seq(40,130)
mu_prior <- dnorm(mu_grid,mu0,sigma0)
mu_posterior <- dnorm(mu_grid,mupost.mean,mupost.sd)

mu_res <- tibble::tibble(mu_grid, mu_prior,mu_posterior) %>% 
  tidyr::pivot_longer(-mu_grid,
                      names_to = "distribution",
                      values_to = "value") %>% 
  dplyr::mutate(distribution = factor(distribution, levels = c("mu_prior", "mu_posterior")))

# plot prior and posterior
ggplot(mu_res, aes(x = mu_grid, y = value, colour = distribution)) +
  geom_line() +
  labs(colour = "") +
  facet_wrap(~distribution, scales = "free_y") +
  ylab("Density") +
  xlab(expression(mu)) +
  facet_wrap(~distribution, scales = "free_y") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```
  
  - Result: $\hat{\mu} =$ `r round(mupost.mean,2)`, 95\% CI: (`r round(lwr,2)`, `r round(upr,2)`)


## Normal distribution with known mean

\footnotesize
  - Assume $\mu$ is known where $\mu = 86.79$
  
- Usually work with precision i.e., $\tau = 1/\sigma^2$

 - Specify a prior for $\tau$

    - Popular prior for the precision of a normal distribution is a gamma prior
e.g., $\tau \sim Gamma(a, b)$ where $E[\tau] = \frac{a}{b}$ and $Var[\tau] = \frac{a}{b^2}$

    - $p(\tau|a,b) = \frac{b^a}{\Gamma(a)}\tau^{a-1}e^{-b\tau}$ for $\tau >0$ and $a,b > 0$

- Use Bayes' rule to obtain the posterior distribution

$p(\tau|y) \propto p(y|\tau)p(\tau)$

- The posterior will also  be a gamma distribution

 $\tau|y \sim Gamma \bigg(a + n/2, b + 1/2\sum_{i=1}^n (y_i - \mu)^2\bigg)$


## Prior vs Posterior for $\tau$
\footnotesize
 - Choose parameter values for the prior distribution for $\tau$, e.g., $a = 0.1$, $b = 0.1$


```{r, fig.width=7, fig.height=2}
#-------------------
# exact solutions
# prior and posterior for tau, conditional on mu (assuming mu = ybar)
mu.y <- 87

# prior parameters
a <- 0.1
b <- 0.1

# Posterior of tau
post.a = a + n/2
post.b = b + 1/2*(sum((y-mu.y)^2))

# Credible intervals
lwr <- qgamma(0.025, post.a, post.b)
upr <- qgamma(0.975, post.a, post.b)

# Set up for plotting
tau_grid <- seq(0,3,by = 0.01)
tau_prior <- dgamma(tau_grid,a,b)
tau_posterior <- dgamma(tau_grid,post.a,post.b)

tau_res <- tibble::tibble(tau_grid, tau_prior,tau_posterior) %>% 
  tidyr::pivot_longer(-tau_grid,
                      names_to = "distribution",
                      values_to = "value") %>% 
  dplyr::mutate(distribution = factor(distribution, levels = c("tau_prior", "tau_posterior")))

# sample from the posterior for tau
tau_samps <- rgamma(1000,post.a,post.b)

# get sigma samples and summaries
sigma_samps <- sqrt(1/tau_samps)

CI <- quantile(sigma_samps, probs = c(0.025,0.5,0.975))

ggplot(tau_res, aes(x = tau_grid, y = value, colour = distribution)) +
  geom_line() +
  labs(colour = "") +
  ylab("Density")+
  xlab(expression(tau))+
  facet_wrap(~distribution, scales = "free_y") +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```
 
 - Result: Through simulation from the posterior for $\sigma$ we find $\hat{\sigma} = 20.23$, 95\% CI: (18.96, 21.63)
 
 
